Evolutionary demography and the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites

### 2020-12-01 doc/SA-Hermaphrodites-wDemography.tex
Draft of Mimulus methods
Added first draft of basic methods for parameterizing the model
using empirical data for M. guttatus

%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Demographic rate estimates for {\itshape M. guttatus}} \label{sec:MguttMethods}

To illustrate the use of our model with empirical data, and see whether demographic rates observed in natural populations are likely to support balanced SA polymorphisms, we used publicly available demographic and fitness data for natural populations of the hermaphroditic flowering plant {\itshape Mimulus guttatus} ({\itshape Erythranthe guttata}). {\itshape M.~guttatus}

>>>>

%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Demographic and fitness data for {\itshape M. guttatus}} \label{sec:MguttMethods}

As an illustrative example of how our model can be used to explore whether demographic rates observed in natural populations appear likely to support balanced SA polymorphisms, we parameterized our model using empirically estimated demographic rates and fitness data for natural populations of the hermaphroditic flowering plant {\itshape Mimulus guttatus} (Scrophulariaceae; now known as {\itshape Erythranthe guttata}). {\itshape M.~guttatus} is an herbaceous, self-compatible wildflower native to western North America that exhibits remarkable among-population variation in numerous life-history and reproductive traits including selfing rates, inbreeding depression, floral morphology, and annual-to-perennial life-history \citep[e.g.,][]{RitlandGanders1987, Ritland1990, Willis1993, Willis1999a, Willis1999b, WuWillis2008}. Moreover, detailed demographic studies have been conducted on multiple populations of {\itshape M.~guttatus}, with demographic matrix data made publicly available on the demographic database COMPADRE \citep{CompadreDB2020}. Below, we briefly outline how we parameterized our model using the available data; full details are provided in \hl{Appendix B.x}.

We used extensive demographic rate data reported for the 'Eagle Meadows' population in Stanislaus National Forest (California, USA) studied by \citet{PetersonEtAl2016}. This study entailed a large common garden experiment which took place at the 'Eagle Meadows' site. We used demographic rates estimated for this population grown in its native habitat (i.e., we did not use data for non-local populations grown at Eagle Meadows). \citet{PetersonEtAl2016} collected data for numerous life-history stages and transition rates; for the Eagle Meadows population these estimates were as follows: seed bank survival ($D = 0.534$), seed germination rate ($G = 0.469$), flower production ($F = 0.64$), ovules per flower ($O = 614)$, seedling recruits proportional to clonal rosette recruits ($A = 6.7e-4$), overwinter survival ($S = 0.179$), rosette production ($R = 8.71$) (see corrected Tables.~1 and S2 in \citealt{PetersonEtAl2017}). These data were also available through the Compadre demographic matrix database \citep{CompadreDB2020}. The resulting transition matrices for this population involved three life-history stages ($\omega = 3$; seed, seedling, and rosette), and individual elements of the transition matrix ($\tilde{\mbf{A}}$) were calculated as products of the above rates (see Matrix 1 in \citealt{PetersonEtAl2016}).

Unfortunately, estimates of selfing rates and inbreeding depression are not also available for the Eagle Meadows population of {\itshape M.~guttatus}. However, selfing rate estimates are available for a variety of other populations, and vary in magnitude from near complete outcrossing to predominant selfing ($C \approx 0$ to $0.75$; \citealt{RitlandGanders1987, Ritland1990, Willis1999b}). Additionally, estimates of inbreeding depression at several of the life-history stages/fitness components that were included in the data of \citet{PetersonEtAl2016} are available for two intensively studied populations in the Cascade Mountains of Oregon (Iron Mountain and Cone Peak) \citet{Willis1993, Willis1999a, Willis1999b}. Using the data provided in \citet{Willis1993}, we estimated the proportional decrease due to inbreeding depression in seed germination rate ($\delta_{G} = 0.085$), flower number ($\delta_{F} = 0.2$), and overwinter survival ($\delta_{S} = 0.38$).

Using these combined demographic rates, selfing rate, and inbreeding depression estimates, we constructed a corresponding stage $\times$ genotype mendelian matrix model with a single SA locus affecting female and male fertility (as described above). Using this empirically parameterized model, we are able to make predictions about the scope for demographically viable SA polymorphism at a single locus in a hypothetical population with the same demographic rates as observed in the Eagle Meadows population, for a range of selfing and inbreeding depression rates observed in other natural populations. However, this should not be confused with an explicit prediction for the likelihood of SA polymorphism in the Eagle Meadows population.

Interestingly, a polymorphic chromosomal inversion (Inv6) with apparently SA fitness effects has been identified in the Iron Mountain population of {\itshape M.~guttatus} \citep{LeeKelly2015}. Inv6 segregates at moderate frequency (about $8\%$), and carriers suffer $\approx 30 \%$ loss in pollen viability, but also increased flower production which varies among years. The genetic basis of the fitness effects caused by Inv6 are not yet known, but the net result is a "supergene" with uniquely strong fitness effects that segregates as a single diallelic locus. We estimated dominance and selection coefficients for the effect of Inv6 on pollen viability and flower production from greenhouse crossing experiments using mapping poulations and field estimates. Across four mapping populations, the effect of Inv6 on pollen viability was partially recessive ($h_m \approx 0.35$). Given this average dominance, the selection coefficient in the field is approximately $s_m = 0.88$. It was not possible to estimate the average dominance of the effect of Inv6 on flower production, so we conservatively assume that it was additive ($h_f = 0.5$). Based this assumption, the selection coefficient Inv6 on flower production in the field was $s_f = 0.384$ in 2012, and $s_f = 0.232$ in 2013. As a final proof-of-concept test of our empirically parameterized model \citep[e.g.,][]{Servedio2014}, we asked whether, given the above biologically grounded fitness effect estimates, Inv6 appears to fall in demographically viable SA polymorphic parameter space, as might be expected given its observed frequency in the Iron Mountain population.
